Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Mechanical Perturbations

### 2024-05-31 main.tex
Removed bib url due to format issue.

\section{Introduction}
%
Single-actor bicycle crashes are associated with a suprisingly large number of
reported injuries~\citep{Wegman2024}. At low speeds, from start up to typical
cruising speeds, bicycles themselves are not self-stable and can be challenging
for the rider to balance. Low speed crashes may be reduced if the bicycle was
self-stable at these speeds, possibly relieving the rider of a necessary
control activity. Bicycles can be mechanically modified to lower the speeds at
which they are self-stable~\citep{Astrom2005} but since the 1980s it has been
known that applying a motor acutated steering torque proportional to the roll
rate can stabilize a single track vehicle at very low speeds. If this automatic
control of steering can stabilize a bicycle, it may reduce the control required
by the rider to successfully manage balancing tasks. We have developed such a
balance assisting bicycle and hypothesize that it helps the rider in situations
they are likely to fall.

In the early years of developments automatic control\citep{Whipple1899}
remarkably realized and showed that roll motion feedback can stabilize the a
bicycle. Automatic roll stabilization of single track vehicles began in earnest
after predictive motorcycle models were developed and refined in throughout the
1970s. \Citet{vanZytveld1975} was influenced by Whipple's work and seems to be
the first \todo{Check if a Japanese researcher was first} to attempt to
robotically stabilize a small motorbike with a controlled inverted pendulum
that mimicked rider lean, but he was not successful in demonstrating what his
theoretic control model correctly predicted. In his model, he recognized that
feedback of the vehicle roll angle and angular rate was essential to stabilize
the vehicle. It was not until the mid 1980s that Ruijs and Pacjecka
successfully demonstrated an automatically balanced
motorcycle~\citep{Ruijs1986} and they did so with a steering motor. Ruijs and
Pacejka clearly showed that steer torque driven by roll angle feedback
stabilizes the capsize mode, roll angular rate feedback stabilizes the weave
mode, and steer angular rate feedback stabilizes the wobble mode. They also
propose how the gains must change with respect to vehicle speed for favorable
control at all speeds. This roll motion feedback is the simplest controller
that can stabilize a single track vehicle above a minimum speed when not
concerned with wobble instabilities, which are not prominent in bicycles at low
speeds. Ruijs and Pacejka's work was not particularly concerned with low speed
stability and their vehicle was fully automatic, no human rider was involved.
Many more automatically balanced single track vehicles have been demonstrated
over the last 40 years, but none have demonstrated that increasing low speed
stability can assist in balancing and possibly reduce single actor falls. Most
of these robotic bicycles and motorcycles designers did not intend for a human
rider to also control the stabilized vehicle. Nevertheless, an automatically
stabilized bicycle can be controlled by a human rider if the motor controlled
steer torque and the rider applied steer torque act on the steer in parallel.
The effect of this automatic control gives the ability to change the dynamics,
up to some limits, of the human controlled plant.In our prior study, we
demonstrate reduced motion during distractions~\cite{Alizadehsaravi2023} due to
the balance assist but researchers with a similar vehicle recently showed rider
disatisfction with the stabilization~\cite{Hanakam2023}.

>>>>

\section{Introduction}
%
Single-actor bicycle crashes are associated with a suprisingly large number of
reported injuries~\citep{Wegman2024}. At low speeds, from start up to typical
cruising speeds, bicycles themselves are not self-stable and can be challenging
for the rider to balance. Low speed crashes may be reduced if the bicycle was
self-stable at these speeds by  relieving the rider of a necessary control
activity. Bicycles can be mechanically modified to lower the speeds at which
they are self-stable~\citep{Astrom2005} but since the 1980s, it has been known
that applying a motor acutated steering torque proportional to the roll rate
can stabilize a single track vehicle at very low speeds. If this automatic
control of steering can stabilize a bicycle, it may reduce the control required
by the rider to successfully manage balancing tasks. We have developed such a
balance assisting bicycle and hypothesize that it helps the rider in situations
they are likely to fall.

In the early years of developments automatic control, \citep{Whipple1899}
remarkably realized and showed that roll motion feedback can stabilize the a
bicycle. Much later, automatic roll stabilization of single track vehicles
began in earnest after predictive motorcycle models were developed and refined
in throughout the 1970s. \Citet{vanZytveld1975} was influenced by Whipple's
work and seems to be the first \todo{Check if a Japanese researcher was first}
to attempt to robotically stabilize a small motorbike with a controlled
inverted pendulum that mimicked rider lean, but he was not successful in
demonstrating what his theoretic control model correctly predicted. In his
model, he recognized that feedback of the vehicle roll angle and angular rate
was essential to stabilize the vehicle. It was not until the mid 1980s that
Ruijs and Pacjecka successfully demonstrated an automatically balanced
motorcycle~\citep{Ruijs1986} and they did so with a steering motor. Ruijs and
Pacejka clearly showed that steer torque driven by roll angle feedback
stabilizes the capsize mode, roll angular rate feedback stabilizes the weave
mode, and steer angular rate feedback stabilizes the wobble mode. They also
propose how the gains must change with respect to vehicle speed for favorable
control at all speeds. This roll motion feedback is the simplest controller
that can stabilize a single track vehicle above a minimum speed when one is not
concerned with wobble instabilities, which are not prominent in bicycles at low
speeds. Ruijs and Pacejka's work was not particularly concerned with low speed
stability and their vehicle was fully automatic, no human rider was involved.
Many more automatically balanced single track vehicles have been demonstrated
over the last 40 years, but none have demonstrated that increasing low speed
stability can assist in balancing and possibly reduce single actor falls. Most
of these robotic bicycle and motorcycle designers did not intend for a human
rider to also control the stabilized vehicle. Nevertheless, an automatically
stabilized bicycle can be controlled by a human rider if the motor controlled
steer torque and the rider applied steer torque act on the steer in parallel.
The effect of this automatic control gives the ability to change the dynamics,
up to some limits, of the human controlled plant. In our prior study, we
demonstrate reduced motion during distractions~\cite{Alizadehsaravi2023} due to
the balance assist but conversely researchers with a similar vehicle recently
showed rider disatisfction with the stabilization~\cite{Hanakam2023}.
